---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Use case 1 {#use-case-1}

## Introduction
In the preceding chapters we have looked at four methods for supervised learning: Nearest Neighbours (chapter \@ref(nearest-neighbours)), Support Vector Machines (chapter \@ref(svm)), Decision Trees (chapter \@ref(decision-trees)) and Random Forests (chapter \@ref(decision-trees)). In this chapter we will consider how we choose the best method for a particular data set.

## Problem: automated detection of malaria

(@Park2016)

[https://doi.org/10.1371/journal.pone.0163045](https://doi.org/10.1371/journal.pone.0163045)

Variables:

* Max optical path length (OPL)
* Mean OPL
* Median OPL
* Optical volume
* Minor axis length
* Major axis length
* Perimeter
* Area
* Elongation
* Equivalent diameter
* Eccentricity
* Standard deviation OPL
* Variance OPL
* Interquartile range OPL
* Skewness
* Kurtosis
* Mean symmetry
* Min symmetry
* Maximum gradient
* Mean gradient
* Centroid vs centre of mass
* Upper quartile mean OPL
* Upper quartile mean gradient

## Challenges



## Getting started

### Load data
```{r}
load("data/malaria/malaria.RData")
```

Inspect objects that have been loaded into R session
```{r}
ls()
class(morphology)
dim(morphology)
names(morphology)
class(infectionStatus)
summary(as.factor(infectionStatus))
class(stage)
summary(as.factor(stage))
```

**morphology** is a data.frame of the morphological parameters from the red blood cells.

**infectionStatus** a factor identifying a red blood cell as either infected or uninfected.

**stage** a factor giving the stage of infection: early trophozoite, late trophozoite, schizont or uninfected.


## Solutions
We have provided worked examples in appendix \@ref(use-case-1-solutions). Can you come up with better solutions?

