# Linear models and matrix algebra {#linear-models}

<!-- Sudhakaran -->

## Linear models 

We will start with simple linear functions.

**First Example**
Height and Weight correlation

```{r out.width='80%', fig.asp=0.744, fig.align='center', echo=T}
people <- read.csv("data/Linear_models/people.csv", header=F)
caption = 'Height and weight correlation of people in USA.'
plot(people$V2 ~ people$V3, ylab="weight", xlab="height", xlim=c(0,200), ylim=c(0,100))

```


```{r out.width='80%', fig.asp=0.744, fig.align='center', echo=T}
knitr::kable(
  head(people[, 1:3], 15), booktabs = TRUE,
  caption = 'A table of height and weight correlation.'
)
```



**Hypothesis** 

$$
h_{\theta}(x) = \theta_{(0)} + \theta_{(1)}x
$$

How do we evaluate?
$$\theta_{(i's)}$$
Let us assume $\theta_{(0)}$ = 25 and $\theta_{(1)}$ = 0

or 

Let us assume $\theta_{(0)}$ = 0 and $\theta_{(1)}$ = -100


Our hope is that the hypothesis $h(x)$ accounts for all of the data with minimal error. 

Mathematically: 

We are tying to minimise $\theta_{(0)}$ and $\theta_{(1)}$

or **minimisation** of 

$$
(h_{\theta}(x) - y)^2
$$
When done for all elements in the matrix it is called the **Cost Function**. 

$$
\begin{equation*} 
 1/2n\sum_{i=1}^{n} (h_{\theta}(x^i) - y^i)^2
\end{equation*} 
$$
where $h_{\theta}(x^i)$ = $\theta_{0} + \theta_{1x^i}$

The **Cost Function** 
$$
\begin{equation*} 
 CF(\theta_{0},\theta_{1}) = 1/2n\sum_{i=1}^{n} (h_{\theta}(x^i) - y^i)^2
\end{equation*} 
$$
 **Cost Function** is also called the **Squard Error Function**
 
Our training data set is a scatter plot in the x-y plane and the straight line or hypothesis defined by $hθ(x)$ has to pass through most of these points and the best possible line will have the least average squared vertical distances from the line. 

When that happens the value of $CF(\theta_{0},\theta_{1})$ will be 0. 

```{r out.width='80%', fig.asp=0.744, fig.align='center', echo=T}
people <- read.csv("data/Linear_models/people.csv", header=F)
caption = 'Height and weight correlation of people in USA.'
plot(people$V2 ~ people$V3, ylab="weight", xlab="height", xlim=c(0,200), ylim=c(0,100))
fit <- lm(people$V2 ~ people$V3)
abline(fit$coef,lwd=2)
b <- round(fit$coef,4)
text(10, 80, paste("y =", b[1], "+", b[2], "x"), adj=c(0,0.5))

```

**Second Example**
Falling of an object

```{r}
library(UsingR)
library(rafalib)
set.seed(1)
g <- 9.8 ##meters per second
n <- 25
tt <- seq(0,3.4,len=n) ##time in secs, note: we use tt because t is a base function
d <- 56.67  - 0.5*g*tt^2 + rnorm(n,sd=1) ##meters

```

```{r}
library(rafalib)
mypar()
plot(tt,d,ylab="Distance in meters",xlab="Time in seconds")
```

This data looks like it might follow the equation

$$ Y_i = \theta_0 + \theta_1 x_i + \theta_2 x_i^2 + \varepsilon_i, i=1,\dots,n $$
With $Y_i$ representing location, $x_i$ representing the time, and $\varepsilon_i$ accounting for measurement error. This is still a linear model because it is a linear combination of known quantities (the $x$'s) referred to as predictors or covariates and unknown parameters (the $\theta$'s).

**Third Example**
Father and son height correlation

```{r}
data(father.son,package="UsingR")
x=father.son$fheight
y=father.son$sheight
plot(x,y,xlab="Father's height",ylab="Son's height")
```
$$ Y_i = \theta_0 + \theta_1 x_i + \varepsilon_i, i=1,\dots,N $$
This is also a linear model with $x_i$ and $Y_i$, the father and son heights respectively, for the $i$-th pair and $\varepsilon_i$ a term to account for the extra variability. Here we think of the fathers' heights as the predictor and being fixed (not random) so we use lower case. Measurement error alone can't explain all the variability seen in $\varepsilon_i$. This makes sense as there are other variables not in the model, for example, mothers' heights, genetic randomness, and environmental factors

**Fourth Example**
Mouse diet data 

Here we read-in mouse body weight data from mice that were fed two different diets: high fat and control (chow). We have a random sample of 12 mice for each. We are interested in determining if the diet has an effect on weight. 

```{r}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv"
filename <- "femaleMiceWeights.csv"
if (!file.exists(filename)) download(url,destfile=filename)

```

```{r}
dat <- read.csv("femaleMiceWeights.csv")
mypar(1,1)
stripchart(Bodyweight~Diet,data=dat,vertical=TRUE,method="jitter",pch=1,main="Mice weights")
```

We can estimate the difference in average weight between populations using a linear model of the form. 

$$ Y_i = \theta_0 + \theta_1 x_{i} + \varepsilon_i$$
with $\theta_0$ the chow diet average weight, $\theta_i$ the difference between averages, $x_i = 1$ when mouse $i$ gets the high fat (hf) diet, $x_i = 0$ when it gets the chow diet, and $\varepsilon_i$ explains the differences between mice of the same population.

**Linear models in general**

We have seen four very different examples in which linear models can be used. A general model that encompasses all of the above examples is the following:

$$ h_{\theta}(x) = \theta_0 + \theta_1 x_{i,1} + \theta_2 x_{i,2} + \dots + \theta_2 x_{i,p} + \varepsilon_i, i=1,\dots,n $$

$$ h_{\theta}(x) = \theta_0 + \sum_{j=1}^p \theta_j x_{i,j} + \varepsilon_i, i=1,\dots,n $$

Note that we have a general number of predictors $p$. Matrix algebra provides a compact language and mathematical framework to compute and make derivations with any linear model that fits into the above framework.

Therefore most inear models are typically described in matrix algebra framework.

## Matrix algebra

The function matrix creates matrices
**matrix (data, nrow, ncol, byrow)**

Matrix fills values by columns
```{r}
seq1 <- seq(1:6)
m1 <- matrix(seq1, 2)
m1
```

You can also fill it by rows 
```{r}
m2 <- matrix(seq1, 2, byrow = T)
m2
```

Creating a matrix of 20 numbers from a standard normal distribution
```{r}
m3 <- matrix(rnorm(20), 4)
m3
```

appending a vector to a matrix
```{r}
v1 <- c(1, 5, 7, 8)
m4 <- cbind(m3, v1)
m4
```

```{r}
v2 <- c(1:6)
m5 <- rbind(m4, v2)
m5
```

Determining the dimension of a matrix

```{r}
dim(m5)
```

```{r}
m6 <- matrix(1:6, 2)
m6
    
m7 <- matrix(c(rep(1, 3), rep(2, 3)), 2, byrow = T)
m7
     
```

Matrix addition

```{r}
m6+m7
```
Matrix subtraction

```{r}
m6-m7
```

Matrix inverse

$$
\begin{align}
X &= \begin{bmatrix}a& b \\
c & d\\
\end{bmatrix}\\\\
X^{-1} &= \dfrac{1}{(ad-bc)}\begin{bmatrix} d& -b\\
-c & a\\
\end{bmatrix}
\end{align}
$$
```{r}
m8 <- matrix(1:4, 2)
m8
```

R function for inv matrix

```{r}
solve(m8)
```

Matrix transpose
$$
\begin{align}
X &= \begin{bmatrix}a& b \\
c & d\\
\end{bmatrix}\\\\
X^{T} &= \begin{bmatrix} a& c\\
b & d\\
\end{bmatrix}
\end{align}
$$
R function for matrix transpose

```{r}
t(m7)
```

Matrix multiplication

Element-wise multiplication

```{r}
m6 * m7
```

Cross product
```{r}
m6 %*% t(m7)
```


Matrices are not commutative: A∗B≠B∗A
Matrices are associative: (A∗B)∗C=A∗(B∗C)

Identity matrix (I)

The identity matrix, when multiplied by any matrix of the same dimensions, results in the original matrix. It's just like multiplying numbers by 1. The identity matrix simply has 1's on the diagonal (upper left to lower right diagonal) and 0's elsewhere.

$$
\begin{bmatrix}1& 0& 0 \\
0& 1& 0 \\
0& 0& 1
\end{bmatrix}\\\\
$$
$AA^{-1} = I$

**Estimating parameters** $\theta$

For the models above to be useful we have to estimate the unknown $\theta$s. In the second example, we want to describe a physical process for which we can't have unknown parameters. In the third example, we better understand inheritance by estimating how much, on average, the father's height affects the son's height. In the fourth example, we want to determine if there is in fact a difference: if $\theta_1 \neq 0$.

As explained above, we have to find the values that minimize the distance of the fitted model to the data. We come back to **Cost Function**. 


$$ \sum_{i=1}^n \left( Y_i - \left(\theta_0 + \sum_{j=1}^p \theta_j x_{i,j}\right)\right)^2 $$

Once we find the minimum, we will call the values the least squares estimates (LSE) and denote them with $\hat{\theta}$. The quantity obtained when evaluating the least squares equation at the estimates is called the residual sum of squares (RSS). Since all these quantities depend on $Y$, they are random variables. The $\hat{\theta}$ s are random variables and we will eventually perform inference on them.

What actually happens when we invoke lm?
Inside of lm, we will form a design matrix $\mathbf{X}$ and calculate the Cost function: $\boldsymbol{\beta}$, which minimizes the sum of squares. The formula for this solution is:

$$ \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{Y} $$
We can calculate this in R using matrix multiplication operator %*%, the inverse function solve, and the transpose function t.


Getting back to the mice example

```{r}
set.seed(1) #same jitter in stripchart
```

```{r}
dat <- read.csv("femaleMiceWeights.csv") ##previously downloaded
stripchart(dat$Bodyweight ~ dat$Diet, vertical=TRUE, method="jitter",
           main="Bodyweight over Diet")
```
We can see that the high fat diet group appears to have higher weights on average, although there is overlap between the two samples.

```{r}
levels(dat$Diet)
X <- model.matrix(~ Diet, data=dat)
head(X)
```
```{r}
Y <- dat$Bodyweight
X <- model.matrix(~ Diet, data=dat)
solve(t(X) %*% X) %*% t(X) %*% Y
```

These coefficients are the average of the control group and the difference of the averages:

```{r}
s <- split(dat$Bodyweight, dat$Diet)
mean(s[["chow"]])
mean(s[["hf"]]) - mean(s[["chow"]])
```

Finally, we use lm to run the linear model:

```{r}
fit <- lm(Bodyweight ~ Diet, data=dat)
summary(fit)
(coefs <- coef(fit))
```

The following plot provides a visualization of the meaning of the coefficients with colored arrows.

```{r}
stripchart(dat$Bodyweight ~ dat$Diet, vertical=TRUE, method="jitter",
           main="Bodyweight over Diet", ylim=c(0,40), xlim=c(0,3))
a <- -0.25
lgth <- .1
library(RColorBrewer)
cols <- brewer.pal(3,"Dark2")
abline(h=0)
arrows(1+a,0,1+a,coefs[1],lwd=3,col=cols[1],length=lgth)
abline(h=coefs[1],col=cols[1])
arrows(2+a,coefs[1],2+a,coefs[1]+coefs[2],lwd=3,col=cols[2],length=lgth)
abline(h=coefs[1]+coefs[2],col=cols[2])
legend("right",names(coefs),fill=cols,cex=.75,bg="white")
```

Data source: Some of this data and code were obtained from 
https://github.com/genomicsclass/labs/blob/master/matrixalg/matrix_algebra_examples.Rmd

**Excersises**

Fit linear models for example 2 and example 3 using lm function. 

Solutions to exercises can be found in appendix \@ref(solutions-linear-models)
