# Solutions ch. 2 - Dimensionality reduction {#solutions-dimensionality-reduction}

Solutions to exercises of chapter \@ref(dimensionality-reduction).

## Exercise 2.1

We will first run PCA on the data. Recall that the data is already log_2 normalised, with expression values beginning from row 4. Within R we would run:

```{r echo=T}
set.seed(12345)
D <- read.csv(file = "data/PGC_transcriptomics/PGC_transcriptomics.csv", header = TRUE, sep = ",", row.names=1)
genenames <- rownames(D)
genenames <- genenames[4:nrow(D)]
```


```{r echo=T}
pcaresult <- prcomp(t(D[4:nrow(D),1:ncol(D)]), center = TRUE, scale = FALSE)
```

Here we have opted to centre the data, but have not normalised each gene to be zero-mean. This is beacuse we are dealing entirely with gene expression, rather than a variety of variables that may exist on different scales. 

We can extract the positions of individual cells from the \texttt{pcaresult$x} variable. In the snipped below we index the different cells types (ESC, pre-implantation cells, primordial germ cells and somatic cells) for easier plotting. 

```{r echo=T}
y1 <- pcaresult$x[which(D[1,]==-1),1:2] # PCA
y2 <- pcaresult$x[which(D[1,]==0),1:2]  #
y3 <- pcaresult$x[which(D[1,]==1),1:2]  #
y4 <- pcaresult$x[which(D[1,]==2),1:2]  #
```

Finally, we can plot the data as follows:

```{r echo=T}
plot(y1,type="p",col="red",xlim=c(-100, 100),ylim=c(-50, 50))
points(y2,type="p",col="black")
points(y3,type="p",col="blue")
points(y4,type="p",col="green")
legend(-95, 50, legend=c("ESC", "preimp", "PGC", "soma"), col=c("red", "black", "blue", "green"), pch="o", bty="n", cex=0.8)
```

From the plot, we can see PCA has done a reasonable job of separating out various cells. For example, a cluster of PGCs appears at the top of the plot, with somatic cells towards the lower right hand side. Pre-implantation embryos and ESCs appear to cluster together: perhaps this is not surprising as ESCs are derived from blastocyst cells. Loosely, we can interpret PC1 as dividing pre-implantation cells from somatic cells, with PC2 separating out PGCs.

## Exercise 2.2. 

In the previous exercise we used PCA to reduce the dimensionality of our data from thousands of genes down to two principle components. By eye, PCA appeared to do a reasonable job separating out different cell types. A useful next step might therefore be to perform clustering on the reduced dimensional space, which can be done using:

```{r echo=T}
clust <- kmeans(pcaresult$x[,1:2], 4, iter.max = 1000)
```

We can now compare the cluster assignment to the known cell types:

```{r echo=T}
Labels <- vector("character", ncol(D))
Labels[which(D[1,]==-1)] = "ESC"
Labels[which(D[1,]==0)] = "preimp"
Labels[which(D[1,]==1)] = "PGC"
Labels[which(D[1,]==2)] = "soma"

clusterresults <- rbind(Labels,clust$cluster)
```

We note that, in general PGCs fall into one or more separate clusters, with soma also separating out well. ESCs and pre-implantation tend to fall into identical clusters. We can take a look at what cell types fall into a specific cluster:

```{r echo=T}
clusterresults[1,which(clusterresults[2,]==1)]
clusterresults[1,which(clusterresults[2,]==2)]
clusterresults[1,which(clusterresults[2,]==3)]
clusterresults[1,which(clusterresults[2,]==4)]
```

## Exercise 2.3.

In our previous section we identified clusters associated with various groups. In our application cluster 1 was associated primarily with pre-implantation cells, with cluster 3 associated with PGCs. We could therefore empirically look for genes that are differentially expressed. Since we know SOX17 is associated with PGC specification in humans [@irie2015sox17,@tang2015unique] let's first compare the expression levels of SOX17 in the two groups:

```{r echo=T}
t.test(D[which(genenames=="SOX17")+3, which(clusterresults[2,]==1)],D[which(genenames=="SOX17")+3, which(clusterresults[2,]==3)])
```

Typically we won't always know the important genes, but can perform an unbiased analysis by testing all genes.

```{r echo=T}
pvalstore <- vector(mode="numeric", length=length(genenames))
for (i in c(1:length(genenames))){
pvals <- t.test(D[which(genenames==genenames[i])+3, which(clusterresults[2,]==1)],D[which(genenames==genenames[i])+3, which(clusterresults[2,]==3)])
pvalstore[i]  <-  pvals$p.value
}
sortedgenes <- genenames[order(pvalstore)]
```

## Exercise 2.4

Within our example, the original axes of our data have very obvious solutions: the axes represent the expression levels of individual genes. The PCs, however, represent linear combinations of various genes, and do not have obvious interpretations. To find an intuition, we can project the original axes (genes) into the new co-ordinate system. This is stored in \texttt{pcaresult$rotation} variable.

```{r echo=T}
plot(pcaresult$rotation[,1:2],type="n")
text(pcaresult$rotation[,1:2], genenames, cex = .4)
```

Okay, this plot is a little busy, so let's focus in on a particular region. Recall that PGCs seemed to lie towards the upper section of the plot (that is PC2 separated out PGCs from other cell types), so we'll take a look at the top section:

```{r echo=T}
plot(pcaresult$rotation[,1:2],type="n",xlim=c(-0.07, 0.07),ylim=c(0.04, 0.1))
genenames <- rownames(D)
genenames <- genenames[4:nrow(D)]
text(pcaresult$rotation[,1:2], genenames, cex = .4)
```

We now see a number of genes that are potentially associated with PGCs. These include a number of known PGCs, for example, both SOX17 and PRDM1 (which can be found at co-ordinates PC1=0, PC2= 0.04) represent two key specifiers of human PGC fate [@irie2015sox17,@tang2015unique,@kobayashi2017principles]. We further note a number of other key regulators, such as DAZL, have been implicated in germ cell development, with DAZL over expressed ESCs forming spermatogonia-like colonies in a rare instance upon xenotransplantation [@panula2016over].

We can similarly look at regions associated with early embryogenesis by concentrating on the lower half of the plot:

```{r echo=T}
plot(pcaresult$rotation[,1:2],type="n",xlim=c(0.0, 0.07),ylim=c(-0.07, -0.03))
genenames <- rownames(D)
genenames <- genenames[4:nrow(D)]
text(pcaresult$rotation[,1:2], genenames, cex = .4)
```

This appears to identify a number of genes associated with embryogenesis, for example, DPPA3, which encodes for a maternally inherited factor, Stella, required for normal pre-implantation development [@bortvin2004dppa3,@payer2003stella] as well as regulation of transcriptional and endogenous retrovirus programs during maternal-to-zygotic transition [@Huang2017stella].

## Exercise 2.5. 

We can run tSNE using the following command:

```{r echo=T}
library(Rtsne)
set.seed(1)
tsne_model_1 = Rtsne(as.matrix(t(D)), check_duplicates=FALSE, pca=TRUE, perplexity=100, theta=0.5, dims=2)
```

As we did previously, we can plot the results using:

```{r echo=T}
y1 <- tsne_model_1$Y[which(D[1,]==-1),1:2]
y2 <- tsne_model_1$Y[which(D[1,]==0),1:2]
y3 <- tsne_model_1$Y[which(D[1,]==1),1:2]
y4 <- tsne_model_1$Y[which(D[1,]==2),1:2]

plot(y1,type="p",col="red",xlim=c(-20, 20),ylim=c(-20, 20))
points(y2,type="p",col="black")
points(y3,type="p",col="blue")
points(y4,type="p",col="green")
legend(-20, 10, legend=c("ESC", "preimp", "PGC", "soma"), col=c("red", "black", "blue", "green"),pch="o", bty="n", cex=0.8)
```

## Exercise 2.6.

We can plot the expression patterns for pre-implantation embryos:

```{r echo=T}
y2_0 <- tsne_model_1$Y[which(D[1,]==0 & D[3,]==0),1:2]
y2_1 <- tsne_model_1$Y[which(D[1,]==0 & D[3,]==1),1:2]
y2_2 <- tsne_model_1$Y[which(D[1,]==0 & D[3,]==2),1:2]
y2_3 <- tsne_model_1$Y[which(D[1,]==0 & D[3,]==3),1:2]
y2_4 <- tsne_model_1$Y[which(D[1,]==0 & D[3,]==4),1:2]
y2_5 <- tsne_model_1$Y[which(D[1,]==0 & D[3,]==5),1:2]
y2_6 <- tsne_model_1$Y[which(D[1,]==0 & D[3,]==6),1:2]

plot(y2_0,type="p",col="tomato",xlim=c(-10, 10),ylim=c(0, 10))
points(y2_2,type="p",col="tomato")
points(y2_2,type="p",col="tomato1")
points(y2_3,type="p",col="tomato1")
points(y2_4,type="p",col="tomato2")
points(y2_5,type="p",col="tomato3")
points(y2_6,type="p",col="tomato4")
legend(-10, 10, legend=c("Ooc", "Zyg", "2C", "4C","8C","Mor","Blast"), col=c("tomato", "tomato", "tomato1", "tomato1", "tomato2","tomato3","tomato4"),pch="o", bty="n", cex=0.8)
```
