# Solutions ch. 7 - Nearest neighbours {#solutions-nearest-neighbours}

Solutions to exercises of chapter \@ref(nearest-neighbours).

## Exercise 1

Load libraries
```{r echo=T}
library(caret)
library(RColorBrewer)
library(doMC)
library(corrplot)
```

Prepare for parallel processing
```{r echo=T}
registerDoMC(detectCores())
```

Load data
```{r echo=T}
load("data/wheat_seeds/wheat_seeds.Rda")
```

Partition data
```{r echo=T}
set.seed(42)
trainIndex <- createDataPartition(y=variety, times=1, p=0.7, list=F)
varietyTrain <- variety[trainIndex]
morphTrain <- morphometrics[trainIndex,]
varietyTest <- variety[-trainIndex]
morphTest <- morphometrics[-trainIndex,]

summary(varietyTrain)
summary(varietyTest)
```

Data check: zero and near-zero predictors
```{r echo=T}
nzv <- nearZeroVar(morphTrain, saveMetrics=T)
nzv
```

Data check: are all predictors on same scale?
```{r echo=T}
summary(morphTrain)
```

```{r wheatBoxplots, fig.cap='Boxplots of the 7 geometric parameters in the wheat data set',,  out.width='75%', fig.asp=1, fig.align='center', echo=T }
featurePlot(x = morphTrain, 
            y = varietyTrain, 
            plot = "box", 
            ## Pass in options to bwplot() 
            scales = list(y = list(relation="free"),
                          x = list(rot = 90)),  
            layout = c(3,3))
```

Data check: pairwise correlations between predictors
```{r wheatCorrelogram, fig.cap='Correlogram of the wheat seed data set.', out.width='75%', fig.asp=1, fig.align='center', echo=T}
corMat <- cor(morphTrain)
corrplot(corMat, order="hclust", tl.cex=1)
```

```{r echo=T}
highCorr <- findCorrelation(corMat, cutoff=0.75)
length(highCorr)
names(morphTrain)[highCorr]
```

Data check: skewness
```{r wheatDensityPlots, fig.cap='Density plots of the 7 geometric parameters in the wheat data set',,  out.width='75%', fig.asp=1, fig.align='center', echo=T }
featurePlot(x = morphTrain, 
            y = varietyTrain,
            plot = "density", 
            ## Pass in options to xyplot() to 
            ## make it prettier
            scales = list(x = list(relation="free"), 
                          y = list(relation="free")), 
            adjust = 1.5, 
            pch = "|", 
            layout = c(3, 3), 
            auto.key = list(columns = 3))
```
            
Create a 'grid' of values of _k_ for evaluation:
```{r echo=T}
tuneParam <- data.frame(k=seq(1,50,2))
```
            
Generate a list of seeds for reproducibility (optional) based on grid size
```{r echo=T}
set.seed(42)
seeds <- vector(mode = "list", length = 101)
for(i in 1:100) seeds[[i]] <- sample.int(1000, length(tuneParam$k))
seeds[[101]] <- sample.int(1000,1)
```

<!--
Define a pre-processor (named transformations) and transform morphTrain
```{r echo=T}
transformations <- preProcess(morphTrain, 
                              method=c("center", "scale", "corr"),
                              cutoff=0.75)
morphTrainT <- predict(transformations, morphTrain)
```
-->

Set training parameters. In the example in chapter \@ref(nearest-neighbours) pre-processing was performed outside the cross-validation process to save time for the purposes of the demonstration. Here we have a relatively small data set, so we can do pre-processing within each iteration of the cross-validation process. We specify the option  ```preProcOptions=list(cutoff=0.75)``` to set a value for the pairwise correlation coefficient cutoff.
```{r echo=T}
train_ctrl <- trainControl(method="repeatedcv",
                   number = 10,
                   repeats = 10,
                   preProcOptions=list(cutoff=0.75),
                   seeds = seeds)
```

Run training
```{r echo=T}
knnFit <- train(morphTrain, varietyTrain, 
                method="knn",
                preProcess = c("center", "scale", "corr"),
                tuneGrid=tuneParam,
                trControl=train_ctrl)
knnFit
```

Plot cross validation accuracy as a function of _k_
```{r cvAccuracyMorphTrain, fig.cap='Accuracy (repeated cross-validation) as a function of neighbourhood size for the wheat seeds data set.', out.width='100%', fig.asp=0.6, fig.align='center', echo=T}
plot(knnFit)
```

Predict the class (wheat variety) of the observations in the test set.
```{r echo=T}
test_pred <- predict(knnFit, morphTest)
confusionMatrix(test_pred, varietyTest)
```


